Anomaly detection with kernel density estimation on manifolds

### 2022-06-07 paper/paper.Rmd
More on KDE in Sec2

Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut]. There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp]. For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef]. However, these density estimators are based on random samples in the Euclidean space.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.

In general, given a multivariate random sample $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation} 
where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$, we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$. We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing. 
@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.

>>>>

Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut]. There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp]. For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ut] or the data point [sample smoothing estimator; @Terrell1992-ut]. However, these density estimators are based on random samples in the Euclidean space.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data. @Scott1992-cj and @Scott2015-vl discussed a number of techniques for estimating the kernel densities of multivariate data. 
In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
drawn from a density $f$, the kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
\end{equation} 
where $\pmb{H}$ is called the bandwidth matrix and it is a $d\times d$ symmetric positive definite matrix, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}|$ being the determinant of matrix $\pmb{H}$.
Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function. And an average over all data points is taken as the estimated density $\hat{f}$. KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point $\pmb{x}$.

The choice of the the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$. It controls the smoothing across all sample data points. For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined. Then the performance of the density estimate $\hat{f}$ depends on the choice of $\pmb{H}$. When $\pmb{H}$ is too small, the density estimates will apprear to be wiggly. While with $\pmb{H}$ being too large, the density estimates will smooth out some local feathers of $f$. So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details]. For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews]. The generalization to multivariate case could mainly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.

Note that the bandwidth matrix in \@ref(eq:kde) is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed. This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut]. In VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $i$ instead of a fixed $\pmb{H}$ as below,
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation}

For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion. Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.

### 2022-06-07 paper/paper.tex
More on KDE in Sec2

Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}. There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}. For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}. However, these density estimators are based on random samples in the Euclidean space.

@@

In general, given a multivariate random sample \(\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation}
where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})\).
When using the standard gaussian kernel, \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\), we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\). We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.

>>>>

Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}. There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}. For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ut}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]}. However, these density estimators are based on random samples in the Euclidean space.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data. \textcite{Scott1992-cj} and \textcite{Scott2015-vl} discussed a number of techniques for estimating the kernel densities of multivariate data.
In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
drawn from a density \(f\), the kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
\end{equation}
where \(\pmb{H}\) is called the bandwidth matrix and it is a \(d\times d\) symmetric positive definite matrix, \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})\).
When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation}
with \(|\pmb{H}|\) being the determinant of matrix \(\pmb{H}\).
Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function. And an average over all data points is taken as the estimated density \(\hat{f}\). KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point \(\pmb{x}\).

The choice of the the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\). It controls the smoothing across all sample data points. For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined. Then the performance of the density estimate \(\hat{f}\) depends on the choice of \(\pmb{H}\). When \(\pmb{H}\) is too small, the density estimates will apprear to be wiggly. While with \(\pmb{H}\) being too large, the density estimates will smooth out some local feathers of \(f\). So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}. For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}. The generalization to multivariate case could mainly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.

Note that the bandwidth matrix in \eqref{eq:kde} is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed. This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}. In VKDE, the estimator in Equation \eqref{eq:vkde} is similar to that in \eqref{eq:kde} except that the bandwidth is a variable matrix \(\pmb{H}_i\) for each \(i\) instead of a fixed \(\pmb{H}\) as below,
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation}

For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion. Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.
